\documentclass[11pt,a4paper]{article}

% ---- Packages ----------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=0.9in]{geometry}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage[hidelinks,colorlinks=true,linkcolor=black,citecolor=teal,urlcolor=teal]{hyperref}
\usepackage{parskip}

% ---- Color palette -----------------------------------------------------
\definecolor{accent}{HTML}{0F766E}
\definecolor{cool}{HTML}{1E3A8A}
\definecolor{neutral}{HTML}{475569}
\definecolor{neutrallight}{HTML}{F1F5F9}

% ---- Header / Footer ---------------------------------------------------
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small Mamba IoT Anomaly Detection}
\fancyhead[R]{\small Madhav Dogra}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ---- Section formatting (tighter) --------------------------------------
\titleformat{\section}{\normalfont\large\bfseries\color{cool}}{}{0em}{}
\titlespacing*{\section}{0pt}{6pt}{2pt}
\setlength{\parskip}{3pt}
\newcommand{\hd}[1]{\textbf{\textcolor{accent}{#1}}}

% =======================================================================
\begin{document}

\begin{center}
{\Large\bfseries Results --- Multi-Source IoT-23 Flow-Pattern Anomaly Detection} \\[2pt]
{\small Madhav Dogra \quad$\cdot$\quad 2026-06-30 \quad$\cdot$\quad IoT-23 (6 scenarios, multi-source)}
\end{center}
\vspace{-2pt}

\section*{Setup}
\hd{Data.} Seven IoT-23 captures (1-1, 3-1, 20-1, 21-1, 42-1, 34-1, 8-1) $\rightarrow$ per-flow first $K{=}32$ packets $\times$ 13 payload-free features; labels joined from each Zeek \texttt{conn.log.labeled} by 5-tuple. Class-balanced and water-filled across captures into \textbf{31{,}960 flows}, in \textbf{3 classes}: Benign 10{,}000, Scan/Botnet 16{,}000, BruteForce 5{,}960. \emph{Taxonomy:} the short-flow attack families --- DDoS, Recon (port-scan) and Mirai C\&C --- are \emph{inseparable in the first $\sim$6 packets} of payload-free data (see Takeaway), so they are reported as one \textbf{Scan/Botnet} class; IoT-23 \texttt{Attack} (telnet credential guessing) $\rightarrow$ BruteForce. Stratified 70/15/15 (test 4{,}794).\\
\hd{Model.} Mamba encoder ($d{=}128$, 4 layers) $+$ softmax classifier $+$ one-class deep-SVDD anomaly head.\\
\hd{Run.} 3 stages $\times$ 6 epochs, stratified split $+$ $\sqrt{}$-inverse-frequency focal weights, CPU (Apple M3 Pro). \textbf{Training $\approx$1\,h}; inference \textbf{5.0\,ms/flow}. Stage~B train acc $\rightarrow$0.98; Stage~C radius$^2$ 0.0298 @ p99.

\section*{Results}

\begin{minipage}[t]{0.46\textwidth}
\centering
\small
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Accuracy             & \textbf{98.2\,\%} \\
Macro-F1 (3 classes) & \textbf{0.984} \\
Weighted-F1          & 0.982 \\
\addlinespace
Anomaly ROC-AUC      & \textbf{0.935} \\
Anomaly PR-AUC       & 0.959 \\
\addlinespace
Latency p50/p95/p99  & 5.0/6.5/6.9\,ms \\
\bottomrule
\end{tabular}

\vspace{8pt}
{\small\textit{Confusion matrix} (rows true, cols pred)}\\[2pt]
\renewcommand{\arraystretch}{1.1}
\begin{tabular}{@{}l rrr@{}}
\toprule
 & B & S & F \\
\midrule
Benign\ \ \,(B)     & 1433 & 63 & 4 \\
Scan/Botnet (S)     & 16 & 2383 & 1 \\
BruteForce\ (F)     & 0 & 1 & 893 \\
\bottomrule
\end{tabular}
\end{minipage}
\hfill
\begin{minipage}[t]{0.50\textwidth}
\centering
\small
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Class} & \textbf{P} & \textbf{R} & \textbf{F1} & \textbf{N} \\
\midrule
Benign      & 0.989 & 0.955 & 0.972 & 1500 \\
Scan/Botnet & 0.974 & 0.993 & 0.983 & 2400 \\
BruteForce  & 0.994 & 0.999 & \textbf{0.997} & 894 \\
\midrule
Macro avg    & 0.986 & 0.982 & \textbf{0.984} & 4794 \\
Weighted avg & 0.982 & 0.982 & 0.982 & 4794 \\
\bottomrule
\end{tabular}

\vspace{8pt}
\raggedright
{\small
\hd{Takeaway.}
\begin{itemize}[leftmargin=1.1em, itemsep=1pt, topsep=2pt]
  \item \textbf{98.2\,\% accuracy}, macro-F1 \textbf{0.984}, over \textbf{7 captures} --- BruteForce F1 0.997, Benign 0.972, Scan/Botnet 0.983.
  \item \hd{Design finding:} payload-free header features on these short flows separate by flow \emph{structure} --- benign ($\sim$3 pkts), short scan/botnet attacks ($\sim$5 pkts), long brute-force ($\sim$30 pkts) --- \emph{not} by fine attack family. DDoS, Recon and Mirai-C\&C overlap in the first few packets (confirmed: class reweighting see-saws, and $>$99\,\% of these flows never reach 32 packets), so they form one Scan/Botnet class.
  \item Anomaly head ROC-AUC 0.935 on multi-source data (honest; the prior single-capture 0.994 was inflated by a Mirai-dominated scenario).
  \item Residual error is the benign$\leftrightarrow$scan boundary (short benign connections vs.\ short scans).
\end{itemize}}
\end{minipage}

\vspace{4pt}
\noindent\textcolor{neutral}{\footnotesize Multi-source run (6 IoT-23 scenarios). The 3-class taxonomy reflects what payload-free per-packet features can resolve on short flows; separating the attack families further (DDoS vs.\ Recon vs.\ Mirai) would require flow-level metadata --- duration, byte counts, destination-port diversity, connection state.}

\end{document}
